---
title: "Market Segmentation Analysis"
author: "Sirius Ife"
date: "2024-03-11"
output: html_document
---

# Introduction

In this project, we will conduct an unsupervised learning analysis on marketing data to uncover patterns and segments within the customer base. The goal is to gain insights that can inform targeted marketing strategies and enhance business decision-making.

## Objective
- Perform exploratory data analysis (EDA) to understand the dataset.
- Pre-process the data by handling missing values, outliers, and encoding categorical variables.
- Conduct dimensionality reduction using PCA (Principal Component Analysis) to identify important features.
- Apply k-means clustering to segment the customer base.
- Interpret the clusters and derive actionable insights for marketing strategies.

# Load Packages

```{r}
library(tidyverse) 
library(lubridate)
library(corrplot)
library(caret)
library(FactoMineR)
library(factoextra)
library(cluster)
```

# Importing Data

```{r}
customers <- read.delim("marketing_campaign.csv", stringsAsFactors = FALSE)
head(customers)
```

# Exploratory Data Analysis (EDA)

```{r}
# Check dimensions and summary statistics
dim(customers)
summary(customers)

# Remove rows with missing income values
customers <- customers %>% filter(!is.na(Income))

# Visualize distribution of income and year of birth
ggplot(customers, aes(y = Income)) + geom_boxplot() 
ggplot(customers, aes(Year_Birth)) + geom_boxplot()
```

# Pre-Processing Data

```{r}
# Convert Dt_Customer to date format
customers <- customers %>% mutate(Dt_Customer = as.Date(dmy(Dt_Customer)))

# Create age variable from year of birth
customers <- customers %>% mutate(Age = 2024 - Year_Birth)
customers <- customers %>% filter(Age < 90)  # Remove outliers

# Collapse marital status into two categories: Single & Taken
customers <- customers %>% mutate(Marital_Status = ifelse(Marital_Status %in% c("Divorced", "Widow", "Alone", "Absurd", "YOLO"), "Single", "Taken"))

# Collapse education into two categories: Graduate & Non-graduate
customers <- customers %>% mutate(Education = ifelse(Education %in% c("Graduation", "PhD", "Master"), "Graduate", "Non-graduate"))

# Convert categorical variables to factors
customers <- customers %>% mutate(Marital_Status = as.factor(Marital_Status), Education = as.factor(Education))

# Rename features and create Total_spent variable
customers <- customers %>% rename(wines = MntWines, fruits = MntFruits, meat = MntMeatProducts, fish = MntFishProducts, sweet = MntSweetProducts, gold = MntGoldProds )
customers <- customers %>% mutate(Total_spent = wines + fruits + meat + fish + sweet + gold)

# Remove redundant features
customers <- customers %>% select(- ID, - Year_Birth, - Dt_Customer, - Z_CostContact, - Z_Revenue)
```

# Correlation Analysis

```{r}
# Calculate correlation matrix
cust_cor <- cor(customers[,3:17])
corrplot(cust_cor, method = "color", order = "hclust", addCoef.col = "black", number.cex = .6)
```

# Dimensionality Reduction: PCA

```{r}
# Running PCA
customers_pca <- PCA(customers[, c(3, 6:17, 25:26)], graph = FALSE)

# Summary of PCA
summary(customers_pca)

# Plotting contributions of variables
fviz_pca_var(customers_pca, col.var = "contrib", gradient.cols = c("#002bbb", "#bb2e00"), repel = TRUE)

# Plotting top 5 variables with highest contributions
fviz_pca_var(customers_pca, select.var = list(contrib = 5), repel = TRUE)

# Barplotting the contributions of variables
fviz_contrib(customers_pca, choice = "var", axes = 1, top = 5)
```

# Biplots

```{r}
fviz_pca_biplot(customers_pca)
```

# K-Means Clustering

```{r}
# Elbow method to determine optimal k
tot_withinss <- map_dbl(1:10, function(k){
  model <- kmeans(x = customers[, c(3, 6:17, 25:26)], centers = k)
  model$tot.withinss
})

elbow_df <- data.frame(k = 1:10, tot_withinss = tot_withinss)
head(elbow_df)

# Plotting the elbow plot
ggplot(elbow_df, aes(k, tot_withinss)) + geom_line() + scale_x_continuous(breaks = 1:10)

# Silhouette analysis
sil_width <- map_dbl(2:10, function(k){
    model <- pam(customers[, c(3, 6:17, 25:26)], k = k)
    model$silinfo$avg.width
})

sil_df <- data.frame(k = 2:10, sil_width = sil_width)
head(sil_df)

ggplot(sil_df, aes(k, sil_width)) + geom_line() + scale_x_continuous(breaks = 2:10) + labs(y = "Avg sil width")

# K-means clustering with k=2
set.seed(77)
customers_cluster <- kmeans(customers[, c(3, 6:17, 25:26)], centers = 2)

```

